Episodic memory: mental time travel or a quantum `memory wave' function?

### 2021-01-12 paper/main.tex
finished pass through manuscript, updated diffs

One practical way to distinguish content from context representations is to separate out mental (or neural) representations that drift rapidly \citep[content;][]{PolyEtal05a, MannEtal12} or gradually \citep[context;][]{PolyEtal05a, MannEtal11, HowaEtal12, LohnEtal18, FolkEtal18}.  An intuition for why contextual representations might be expected to evolve gradually is laid out by \cite{PolyKaha08}.  Essentially, slowly drifting thoughts may be used to ``index'' more rapidly drifting thoughts that are temporally proximal.  This notion, that our brains maintain parallel mental representations drifting at different timescales, has been well-characterized in a series of elegant fMRI and ECoG studies by Uri Hasson and colleagues~\citep{HassEtal08, LernEtal11, HoneEtal12a, AlyEtal18}.  Subsequent work by the same group has shown that this hierarchy of drifting mental representations plays a central role in how we remember continuous experiences and segment our experiences into discrete events~\citep{BaldEtal17}.  This body of work suggests that different brain regions reflect \textit{temporal receptive windows} that characterize the timescale(s) to which each region is maximally sensitive.  This research complements work on the \textit{temporal window of integration} in the speech and perception literatures~\citep[e.g.,][]{ViemWake91, GautEtal12, vanWEtal07}.  The temporal window of integration refers to the duration over which physically distinct stimuli are treated (by some brain system or network) as a single percept or atomic unit.  This systems-level work mirrors the discovery of \textit{time cells} in the rodent hippocampus that respond preferentially to a specific time relative to a temporal reference (e.g., time elapsed since starting to run on an exercise wheel), whereby each cell appears to integrate incoming information at a different rate~\citep{PastEtal08b, MacDEtal11} and the population activity serves to represent information at a range of timescales~\citep{MauEtal18}. Other work suggests that the lateral entorhinal cortex also plays a role in integrating information across a wide range of timescales~\citep{TsaoEtal18}, supporting precise temporal recall~\citep{MontEtal19}.  In addition to this work suggesting that individual systems or neurons maintain representations at their preferred timescales, a number of studies also report neurons and small populations exhibiting \textit{temporal multiplexing}, whereby information at different timescales is reflected in different features of an individual neuron's (or population's) activity patterns, such as spike timing versus firing rate or local field potential oscillations at different frequencies~\citep{CaruEtal18, KnigEich13, WatrEtal13b, PanzEtal10}.

@@

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal21} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal21} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal21} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people~\citep[also see][]{MortEtal20}, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experiences over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

\subsection*{The \textit{matching problem}}
The trajectories through word embedding space displayed in Figure~\ref{fig:trajectories} demonstrate that topic models are sufficiently expressive and robust to provide one solution to the \textit{matching problem} in memory analysis.  The matching problem refers to discovering an alignment in the labels between experienced and remembered events.  Traditional memory paradigms most commonly studied over the past century~\citep{Kaha12} ask participants to study lists of \textit{items} (often randomly selected words) and then engage in a memory task, such as verbally recalling any words the participant remembers studying, in any order.  Solving the matching problem in such experiments is trivial: if we treat each studied word on a list as an ``event,'' then we can assume that the participant is remembering a word's event when they verbalize that word during their memory test.  In other words, matching entails finding the exact match between a word the participant is verbalizing and that same word on the list that they studied.

For real-world memory phenomena, and in many naturalistic memory paradigms, solving the matching problem is more complex.  People often experience real-world events (and scenes from movies or stories) as rich sensory experiences that derive deep meaning from surrounding moments, other prior experiences, emotions, goals, high-order understanding of the current situation, and so on.  Many of these aspects of our real-world experiences are not immediately verbalizable-- and even when we are able to describe some specific aspects of what we are experiencing, there are typically many possible ways we could have done so (e.g., varying in level of detail, specific wording, objectivity versus subjectivity, contextual details, etc.).  Further, because the way we experience each moment depends in part on our unique prior experiences, even ostensibly shared experiences can \textit{seem} different to each individual.  The way we recount a given experience is often a reflection of our own idiosyncratic experience of the original event, and is colored and distorted in an idiosyncratic way by our memory systems.  This distorted memory must then be verbalized or communicated, which is itself an idiosyncratic (and potentially lossy, in the information theoretic sense) process.  Solving the matching problem under these constraints poses a substantial theoretical and methodological challenge.  Specifically, matching requires building explicit models of the dynamic conceptual content of our experiences and memories, and then estimating when and how that experienced and remembered content aligns.  The approach featured in Figure~\ref{fig:trajectories} uses a word embedding space that is defined by a topic model~\citep{BleiEtal03}.  Each coordinate in this space reflects the conceptual content underlying a set of words, such that two different text passages that describe the same conceptual content (but using different words, levels of detail, etc.) are mapped onto nearby points in the word embedding space.  When these models are sufficiently expressive and robust to solve the matching problem, by modeling our experiences and memories within the same word embedding space, we can see a clear geometric alignment between the trajectories that our experiences (Fig.~\ref{fig:trajectories}A) and memories (Fig.~\ref{fig:trajectories}B) take through word embedding space~\citep[also see][]{HeusEtal21}.

@@

When text embedding models succeed at capturing semantic overlap between different words or text passages, they can provide an elegant means of solving the matching problem.  For example, when two or more individuals describe a common experience, embedding the text of those descriptions in a high dimensional feature space abstracts away the specific word choices and grammatical choices in favor of the descriptions' underlying conceptual content.  Further, geometric differences in those embeddings (e.g., the Euclidean distances between the corresponding vectors) can capture subtle differences in conceptual content that might not have been readily apparent in the original text.  Of course, there are also limits on how accurately specific word embedding models characterize different conceptual aspects of a given text passage, and on how informative or meaningful those differences are from a psychological perspective.  The field of text embedding has advanced rapidly, particularly in recent years, enabling more accurate and informative approaches to characterizing the conceptual content of arbitrary text passages.

@@

The \textit{Neural Probabilistic Language Model}~\citep[NPLM;][]{BengEtal03} was one of the earliest deep learning-based sequence models.  NLPM derives word embeddings for each word in the vocabulary (such that nearby embeddings reflect semantically related words), and then models word sequences according to their meanings.  This model was inspired in part by $n$-gram models~\citep{BrowEtal92}, which estimate the probability of the next word in a sequence given the $n$ words that came before it. Effectively, NLPM attempts to learn arbitrarily long sequences.  In a related approach, \cite{MnihHint09} proposed the \textit{Scalable Hierarchical Distributed Language Model}, which learns sequences of variable length $n$-grams to achieve greater explanatory power.

The next major advances in word embeddings came from attempts to more precisely capture the geometric relations between words.  For example, if each word's semantic meaning is captured by a feature vector, then in principle one could perform vector algebra on different words' embeddings to produce new concepts that combine their meanings.  For example, subtracting the vector for ``man'' from that of ``king'' and then adding the vector for ``woman'' might result in a new feature vector similar to that of ``queen''~\citep{MikoEtal13b}.  \cite{MikoEtal13a} developed the deep learning model \textit{word2vec} to explicitly optimize the accuracy of these vector operations.  Whereas word2vec is a bag of words model, \textit{Global Vectors for Word Representation}~\citep[GloVe;][]{PennEtal14} extends this approach to model each word's local (temporal) context~\citep[also see][]{PeteEtal18}.

@@

Cutting-edge approaches to word embeddings that have achieved the best-to-date performance on modern natural language processing benchmarks are built using enormous deep neural networks often with many billions of parameters.  Because these models are very computationally expensive to fit, an important discovery has been that many layers in these models may be pre-trained without substantially affecting performance, even on new tasks.  One of the first of this new class of model, \textit{Bidirectional Encoder Representations from Transformers}~\citep[BERT;][]{DevlEtal18}, uses a pre-trained deep neural network, whose weights in a single output layer may be fine-tuned to suit a wide range of new tasks.  The latest additions to this class of models are called \textit{Generative Pre-trained Transformers}~\citep[GPT;][]{RadfEtal19, BrowEtal20}.

When the ``surface features'' of our experiences repeat at non-contiguous timepoints, or when one moment overlaps conceptually with others, this overlap can be characterized by text embedding models like those described above.  In turn, this can help explain non-contiguous ``bumps'' in the quantum memory wave function (e.g., Figs.~\ref{fig:qwave}C, \ref{fig:reinstatement}B) and off-diagonal peaks in semantic temporal correlation matrices (e.g., Fig.~\ref{fig:corrmats}).

\subsection*{Extensions to spatial, semantic, affective, and social representations}
The way we navigate through \textit{thought space} (i.e., word embedding spaces; Figs.~\ref{fig:trajectories}, \ref{fig:embedding-models}) differs from how we navigate in physical spatial environments.  First, we cannot teleport in real space, whereas our thoughts can exhibit rapid jumps between different non-contiguous regions of word embedding space.  Second, when two people visit the same fixed physical landmark, they necessarily visit the same physical location.  When two people describe a shared experience, their idiosyncratic thoughts, goals, prior experiences, etc.\ can lead them to process and remember the experience differently, thereby visiting different locations within thought space.  Third, each time we revisit a fixed physical landmark, we (by definition) revisit the same spatial location.  In other words, it is possible to visit the same spatial location multiple times.  However, we cannot revisit our prior experiences in this way.  Rather, we only get to experience each moment once; subsequent ``visits'' to a prior experience must occur through a different medium than the original experience (e.g., our memory, another person’s memory, a physical recording, a written account, etc.).  Fourth, we can revisit different aspects of the same prior experience on different ``visits''-- e.g., when we remember a specific event we can focus on the physical occurrence in one remembering, the emotional occurence in another remembering, the social implications in another remembering, etc.  We can also revisit the same experience in different levels of detail or depth.  There are no deep analogs of these phenomena in spatial navigation.  Finally, \textit{physically} (at a macro scale), we are always located at a single moment and in a single location.  The main argument of the quantum memory wave function framework I propose here is that we may be \textit{mentally} spread over many times or locations simultaneously.  For example, this provides a way for us to integrate information from multiple prior experiences into a single decision or conceptualization, even if those experiences were not temporally contiguous.

>>>>

One practical way to distinguish content from context representations is to separate out mental (or neural) representations that drift rapidly \citep[content;][]{PolyEtal05a, MannEtal12} versus gradually \citep[context;][]{PolyEtal05a, MannEtal11, HowaEtal12, LohnEtal18, FolkEtal18}.  An intuition for why contextual representations might be expected to evolve gradually is laid out by \cite{PolyKaha08}.  Essentially, slowly drifting thoughts may be used to ``index'' more rapidly drifting thoughts that are temporally proximal.  This notion, that our brains maintain parallel mental representations drifting at different timescales, has been well-characterized in a series of elegant fMRI and ECoG studies by Uri Hasson and colleagues~\citep{HassEtal08, LernEtal11, HoneEtal12a, AlyEtal18}.  Subsequent work by the same group has shown that this hierarchy of drifting mental representations plays a central role in how we remember continuous experiences and segment our experiences into discrete events~\citep{BaldEtal17}.  This body of work suggests that different brain regions reflect \textit{temporal receptive windows} that characterize the timescale(s) to which each region is maximally sensitive.  This research complements work on the \textit{temporal window of integration} in the speech and perception literatures~\citep[e.g.,][]{ViemWake91, GautEtal12, vanWEtal07}.  The temporal window of integration refers to the duration over which physically distinct stimuli are treated (by some brain system or network) as a single percept or atomic unit.  This systems-level work mirrors the discovery of \textit{time cells} in the rodent hippocampus that respond preferentially to a specific time relative to a temporal reference (e.g., time elapsed since starting to run on an exercise wheel), whereby each cell appears to integrate incoming information at a different rate~\citep{PastEtal08b, MacDEtal11} and the population activity serves to represent information at a range of timescales~\citep{MauEtal18}. Other work suggests that the lateral entorhinal cortex also plays a role in integrating information across a wide range of timescales~\citep{TsaoEtal18}, supporting precise temporal recall~\citep{MontEtal19}.  In addition to this work suggesting that individual systems or neurons maintain representations at their preferred timescales, a number of studies also report neurons and small populations exhibiting \textit{temporal multiplexing}, whereby information at different timescales is reflected in different features of an individual neuron's (or population's) activity patterns, such as spike timing versus firing rate or local field potential oscillations at different frequencies~\citep{CaruEtal18, KnigEich13, WatrEtal13b, PanzEtal10}.

@@

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized by projecting text embedding representations of the events onto lower-dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal21} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal21} use the geometric shapes of experience trajectories (through text embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal21} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people~\citep[also see][]{MortEtal20}, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experiences over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

\subsection*{The \textit{matching problem}}
The trajectories through text embedding space displayed in Figure~\ref{fig:trajectories} demonstrate that topic models are sufficiently expressive and robust to provide one solution to the \textit{matching problem} in memory analysis.  The matching problem refers to discovering an alignment in the labels between experienced and remembered events.  Traditional memory paradigms most commonly studied over the past century~\citep{Kaha12} ask participants to study lists of \textit{items} (often randomly selected words) and then engage in a memory task, such as verbally recalling any words the participant remembers studying, in any order.  Solving the matching problem in such experiments is trivial: if we treat each studied word on a list as an ``event,'' then we can assume that the participant is remembering a word's event when they verbalize that word during their memory test.  In other words, matching entails finding the exact match between a word the participant is verbalizing and that same word on the list that they studied.

For real-world memory phenomena, and in many naturalistic memory paradigms, solving the matching problem is more complex.  People often experience real-world events (and scenes from movies or stories) as rich sensory experiences that derive deep meaning from surrounding moments, other prior experiences, emotions, goals, high-order understanding of the current situation, and so on.  Many of these aspects of our real-world experiences are not immediately verbalizable-- and even when we are able to describe some specific aspects of what we are experiencing, there are typically many possible ways we could have done so (e.g., varying in level of detail, specific wording, objectivity versus subjectivity, contextual details, etc.).  Further, because the way we experience each moment depends in part on our unique prior experiences, even ostensibly shared experiences can \textit{seem} different to each individual.  The way we recount a given experience is often a reflection of our own idiosyncratic experience of the original event, and is colored and distorted in an idiosyncratic way by our memory systems.  This distorted memory must then be verbalized or communicated, which is itself an idiosyncratic (and potentially lossy, in the information theoretic sense) process.  Solving the matching problem under these constraints poses a substantial theoretical and methodological challenge.  Specifically, matching requires building explicit models of the dynamic conceptual content of our experiences and memories, and then estimating when and how that experienced and remembered content aligns.  The approach featured in Figure~\ref{fig:trajectories} uses a text embedding space that is defined by a topic model~\citep{BleiEtal03}.  Each coordinate in this space reflects the conceptual content underlying a set of words, such that two different text passages that describe the same conceptual content (but using different words, levels of detail, etc.) are mapped onto nearby points in the text embedding space.  When these models are sufficiently expressive and robust to solve the matching problem, by modeling our experiences and memories within the same text embedding space, we can see a clear geometric alignment between the trajectories that our experiences (Fig.~\ref{fig:trajectories}A) and memories (Fig.~\ref{fig:trajectories}B) take through text embedding space~\citep[also see][]{HeusEtal21}.

@@

When text embedding models succeed at capturing semantic overlap between different words or text passages, they can provide an elegant means of solving the matching problem.  For example, when two or more individuals describe a common experience, embedding the text of those descriptions in a high dimensional feature space abstracts away the specific word choices and grammatical choices in favor of the descriptions' underlying conceptual content.  Further, geometric differences in those embeddings (e.g., the Euclidean distances between the corresponding vectors) can capture subtle differences in conceptual content that might not have been readily apparent in the original text.  Of course, there are also limits on how accurately specific text embedding models characterize different conceptual aspects of a given text passage, and on how informative or meaningful those differences are from a psychological perspective~\citep[also see][]{MannKaha12}.  The field of text embedding has advanced rapidly, particularly in recent years, enabling more accurate and informative approaches to characterizing the conceptual content of arbitrary text passages.

@@

The \textit{Neural Probabilistic Language Model}~\citep[NPLM;][]{BengEtal03} was one of the earliest deep learning-based sequence models.  NLPM derives embeddings for each word in the vocabulary (such that nearby embeddings reflect semantically related words), and then models word sequences according to their meanings.  This model was inspired in part by $n$-gram models~\citep{BrowEtal92}, which estimate the probability of the next word in a sequence given the $n$ words that came before it. Effectively, NLPM attempts to learn arbitrarily long sequences.  In a related approach, \cite{MnihHint09} proposed the \textit{Scalable Hierarchical Distributed Language Model}, which learns sequences of variable length $n$-grams to achieve greater explanatory power.

The next major advances in text embeddings came from attempts to more precisely capture the geometric relations between words.  For example, if each word's semantic meaning is captured by a feature vector, then in principle one could perform vector algebra on different words' embeddings to produce new concepts that combine their meanings.  For example, subtracting the vector for ``man'' from that of ``king'' and then adding the vector for ``woman'' might result in a new feature vector similar to that of ``queen''~\citep{MikoEtal13b}.  \cite{MikoEtal13a} developed the deep learning model \textit{word2vec} to explicitly optimize the accuracy of these vector operations.  Whereas word2vec is a bag of words model, \textit{Global Vectors for Word Representation}~\citep[GloVe;][]{PennEtal14} extends this approach to model each word's local (temporal) context~\citep[also see][]{PeteEtal18}.

@@

Cutting-edge approaches to text embeddings that have achieved the best-to-date performance on modern natural language processing benchmarks are built using enormous deep neural networks often with many billions of parameters.  Because these models are very computationally expensive to fit, an important discovery has been that many layers in these models may be pre-trained without substantially affecting performance, even on new tasks.  One of the first of this new class of model, \textit{Bidirectional Encoder Representations from Transformers}~\citep[BERT;][]{DevlEtal18}, uses a pre-trained deep neural network, whose weights in a single output layer may be fine-tuned to suit a wide range of new tasks.  The latest additions to this class of models are called \textit{Generative Pre-trained Transformers}~\citep[GPT;][]{RadfEtal19, BrowEtal20}.

When the ``surface features'' of our experiences repeat at non-contiguous timepoints, or when one moment overlaps conceptually with others, this overlap can be characterized by text embedding models like those described above.  In turn, this can help explain non-contiguous ``bumps'' in the quantum memory wave function (e.g., Figs.~\ref{fig:qwave}C, \ref{fig:reinstatement}B) and off-diagonal peaks in semantic temporal correlation matrices (e.g., Fig.~\ref{fig:corrmats}).  The precision and accuracy with which text embedding models capture the conceptual nuances within a text passage also influence the accuracy and interpretability of the resulting trajectory of an experience or memory through embedding space (e.g., Fig.~\ref{fig:trajectories}).  For example, an accurate but imprecise model might capture the low spatial frequency features of a trajectory's shape (i.e., the high-level details of the experience), whereas an accurate and precise model could also capture the high spatial frequency features (i.e., low-level details).

\subsection*{Extensions to spatial, semantic, affective, and social representations}
The way we navigate through \textit{thought space} (i.e., text embedding spaces; Figs.~\ref{fig:trajectories}, \ref{fig:embedding-models}) differs from how we navigate in physical spatial environments.  First, we cannot teleport in real space, whereas our thoughts can exhibit rapid jumps between different non-contiguous regions of text embedding space.  Second, when two people visit the same fixed physical landmark, they necessarily visit the same physical location.  When two people describe a shared experience, their idiosyncratic thoughts, goals, prior experiences, etc.\ can lead them to process and remember the experience differently, thereby visiting different locations within thought space.  Third, each time we revisit a fixed physical landmark, we (by definition) revisit the same spatial location.  In other words, it is possible to visit the same spatial location multiple times.  However, we cannot revisit our prior experiences in this way.  Rather, we only get to experience each moment once; subsequent ``visits'' to a prior experience must occur through a different medium than the original experience (e.g., our memory, another person’s memory, a physical recording, a written account, etc.).  Fourth, we can revisit different aspects of the same prior experience on different ``visits''-- e.g., when we remember a specific event we can focus on the physical occurrence in one remembering, the emotional occurence in another remembering, the social implications in another remembering, etc.  We can also revisit the same experience in different levels of detail or depth.  There are no deep analogs of these phenomena in spatial navigation.  Finally, \textit{physically} (at a macro scale), we are always located at a single moment and in a single location.  The main argument of the quantum memory wave function framework I propose here is that we may be \textit{mentally} spread over many times or locations simultaneously.  For example, this provides a way for us to integrate information from multiple prior experiences into a single decision or conceptualization, even if those experiences were not temporally contiguous.
